---
title: "analysis of viromes"
author: "Aydin Karatas"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(root.dir = "~/Documents/Research/project_coprolite_viromes/analysis/")
```

## Source Functions

```{r source functions, echo=FALSE}
base_dir <- "~/Documents/Research/project_coprolite_viromes/analysis/"

# function to source R files from a specified directory
source_files_from_dir <- function(directory) {
  full_path <- file.path(base_dir, directory)
  functions_list <- list.files(path = full_path, pattern = "*.R", full.names = TRUE)
  sapply(functions_list, source)
}

# fource functions from different directories
source_files_from_dir("general_R_functions")
source_files_from_dir("clustering")
source_files_from_dir("diff_repres")
source_files_from_dir("predictions")
```

## Define hyperparameters and global variables
```{r pars and vars}
set.seed(seed = 123)

k <- 4
iterations <- 1000
subset_top <- 100
color_range <- 10
star_bounds <- list(list(0.1, 0.9), list(0.025, 0.975), list(0.01, 0.99),
                    list(0.001, 0.999))

violin_dir <- "../project_coprolite_viromes_NOT_CODE/figures/violin"
heat_dir <- "../project_coprolite_viromes_NOT_CODE/figures/heat"
jitter_dir <- "../project_coprolite_viromes_NOT_CODE/figures/jitter"

voilin_vir_ec <- paste0("clustering_vir_ec_", k)
voilin_lifestyle <- "lifestyle"
voilin_phage_ec <- paste0("clustering_phage_ec_", k)
voilin_phage_fam <- paste0("clustering_phage_fam_", k)
voilin_phage_host <- paste0("clustering_phage_host_", k)

heat_vir_ec <- paste0("diff_repres_vir_ec_", subset_top)
heat_vir_ec_no_key <- paste0("diff_repres_vir_ec_no_key", subset_top)
heat_phage_ec <- paste0("diff_repres_phage_ec_", subset_top)
heat_phage_fam <- paste0("diff_repres_phage_fam_", subset_top)
heat_phage_host <- paste0("diff_repres_phage_host_", subset_top)

jitter_vir_classes <- paste0("vir_classes_", subset_top)
jitter_phage_classes <- paste0("phage_classes_", subset_top)
```

## Analysis of viral metabolic ec counts
First examine what the optimal k for k-means should be 
```{r vir ec prep}
# read in raw counts data
vir_ec_counts <- read.csv("~/Documents/Research/project_coprolite_viromes_NOT_CODE/data/vir_phage_ec_counts.csv",
                          header = TRUE, row.names = 1)

# normalize observations within a sample and transform to log( CPM + 1 )
vir_ec_norm <- normalize_data(vir_ec_counts)
vir_ec_t_mat <- scale(t(vir_ec_norm))

# find optimal number of clusters with a wss plot
vir_ec_optimal_wss <- get_optimal_clusters(vir_ec_t_mat, "wss")
vir_ec_optimal_wss
```

Next, perform the bootstrap and permutation k-means; produce a violin plot
```{r vir ec vioin}
# perform kmeans with boostraping and permutations
vir_ec_boot_res <- kmeans_bootstrap(vir_ec_t_mat, k, iterations)
vir_ec_perm_res <- kmeans_permute(vir_ec_t_mat, k, iterations)

# show clustering results by combining bootstrap and permutation res
vir_ec_clus_res <- get_clustering_res(vir_ec_boot_res, vir_ec_perm_res,
                                      star_bounds, iterations)
vir_ec_clus_violin <- create_violin(vir_ec_boot_res, vir_ec_clus_res,
                                    voilin_vir_ec, violin_dir)
```
Produce heatmap of most differential represented metabolic genes
```{r vir ec heat}
# show differential representation results
vir_ec_diff_repres <- get_diff_repres_ec(vir_ec_t_mat)
vir_ec_heatmap <- create_heatmap(vir_ec_t_mat, vir_ec_diff_repres, color_range,
                                 subset_top, heat_vir_ec, heat_dir)
```

Produce jitter plot of enzyme categories
```{r vir ec ezy}
# show ec proportions
vir_ec_median <- get_cat_medians(vir_ec_t_mat[, rownames(vir_ec_diff_repres)[1:subset_top]])
get_diff_repres_class(vir_ec_median)
vir_ec_jitter <- create_jitter(vir_ec_median, c(1:7), jitter_vir_classes,
                               jitter_dir)
```
## Lifestyle prediction 
```{r lifestyle}
# test lifestyle counts
lifestyle_counts <- read.csv("../project_coprolite_viromes_NOT_CODE/data/lifestyle_counts.csv",
                             header = TRUE, row.names = 1)
lifestyle_boot_res <- get_median_lifestyles(lifestyle_counts, iterations)

lifestyle_voilin <- create_lifestyle_voilin(lifestyle_boot_res,
                                            voilin_lifestyle, violin_dir)
lifestyle <- melt(lifestyle_boot_res)
kruskal.test(lifestyle$value, lifestyle$Var2)
```