---
title: "Analysis of Viromes"
author: "Aydin Karatas"
output: html_document
---

## set up functions and working directory
```{r setup, include=FALSE}
base_dir <- "~/Documents/research/project_coprolite_viromes/analysis/"
knitr::opts_chunk$set(root.dir = base_dir)

# source functions from different directories
source_files_from_dir <- function(directory) {
  full_path <- file.path(base_dir, directory)
  functions_list <- list.files(
    path = full_path,
    pattern = "*.R",
    full.names = TRUE
    )
  sapply(functions_list, source)
}
source_files_from_dir("stats")
source_files_from_dir("graphing")
source_files_from_dir("misc")

# define global variables
set.seed(seed = 123)
data_dir <- "~/Documents/research/project_coprolite_viromes_NOT_CODE/data"
figs_dir <- "~/Documents/research/project_coprolite_viromes_NOT_CODE/figures"
max_iter <- 1000
subset_top <- 100
star_bounds <- list(list(0.1, 0.9), list(0.025, 0.975), list(0.01, 0.99),
                    list(0.001, 0.999))
```

## viral contigs vs phage contigs
Generate an image showing the proportion of a sample's virome that is
considered "viral" through a PROKKA-based filter vs. the proportion of a 
sample's virome is phage-like by PhaBOX
```{r get viral contig prop vs phage congtig prop}
vir_vs_phage_counts <- read.csv(
  file = paste0(data_dir, "/viral_vs_phage_prop_all.csv"),
  header = TRUE
)
vir_vs_phage_counts$prokka_perc <-
  vir_vs_phage_counts$prokka_res / vir_vs_phage_counts$total_contigs
vir_vs_phage_counts$phabox_perc <-
  vir_vs_phage_counts$phabox_res / vir_vs_phage_counts$total_contigs

lm = lm(phabox_perc ~ prokka_perc, data = vir_vs_phage_counts)
lm
str(summary(lm))
vir_vs_phage_plot <- 
  ggplot(vir_vs_phage_counts, aes(x = prokka_perc, y = phabox_perc)) +
  geom_point(aes(col = cat)) +
  geom_smooth(method = "lm", col = "blue") + 
  labs(x = "viral contig percentage", y = "phage contig percentage") +
  theme_minimal()
vir_vs_phage_plot
ggsave(filename = paste0(figs_dir, "/vir_vs_phage_plot.svg"),
       plot = vir_vs_phage_plot,
       height = 7,
       width = 14)
```

## lifestyle prediction 
read in lifestyle prediction counts and bootstrap for central tendency
```{r lifestyle boot}
# read in lifestyle counts
lifestyle_counts <- read.csv(
  file = paste0(data_dir, "/lifestyle_counts.csv"),
  header = TRUE,
  row.names = 1
  )
lifestyle_counts["total", ] <- colSums(lifestyle_counts)
lifestyle_counts["temp_to_lyt", ] <- 
  lifestyle_counts["temperate", ] / lifestyle_counts["virulent", ]
lifestyle_counts["temp_to_lyt_+1", ] <- 
  (lifestyle_counts["temperate", ] + 0.01) / (lifestyle_counts["virulent", ] + 0.01)
lifestyle_counts["temp_perc", ] <-
  lifestyle_counts["temperate", ] / lifestyle_counts["total", ]

# bootstrap median ratio of temperate-to-lytic viruses
lifestyle_boot_res <- get_row_median(
  data = lifestyle_counts,
  num_iter = max_iter, 
  row_num = 6
  )

# plot violin plot of lifestyles by category
lifestyle_voilin <- create_cat_voilins(
  data = lifestyle_boot_res, 
  ci_offset = c(0.15, 0.55, 0.55),
  file_name = "lifestyle",
  plot_dir = paste0(figs_dir, "/lifestyle")
  )
lifestyle_voilin

# print stats testing of lifestyles preferences between categories
lifestyle <- melt(lifestyle_boot_res)
kruskal.test(lifestyle$value, lifestyle$Var2)
```

## viral host analysis
determine which predicted hosts are diff. repress
```{r viral host analysis}
# read in raw counts data
host_counts <- read.csv(
  file = paste0(data_dir, "/samples_v_genus.csv"),
  header = TRUE,
  row.names = 1
)

# normalize observation to scaled log( CPM + 1 )
host_norm <- normalize_data(df = host_counts)
host_t_norm <- t(host_norm)

# get differentially represented hosts
host_diff_repress <- get_diff_repres_cols(data = host_t_norm)
```
phyla level analysis
```{r host phyla}
phyla_counts <- read.csv(
  file = paste0(data_dir, "/phyla_counts.csv"),
  header = TRUE,
  row.names = 1)

phyla_norm <- normalize_data(phyla_counts)
phyla_t_norm <- t(phyla_norm)

# get differentially represented hosts
phyla_diff_repress <- get_diff_repres_cols(data = phyla_t_norm)

# generate heatmap of phyla
phyla_t_scale <- scale(phyla_t_norm)
create_phyla_heatmap(
  data = phyla_t_scale,
  signif_res = phyla_diff_repress,
  subset = ncol(phyla_t_scale),
  file_name = "diff_repres_phyla",
  plot_dir = paste0(figs_dir, "/hosts")
  )
```
furmicutes-to-bacterioda analysis
```{r vFB versus time}
time_vFB <- read.csv(
  file = paste0(data_dir, "/time_vs_vFB.csv"),
  header = TRUE,
  row.names = 1)
time_vFB_t_mat <- t(time_vFB)

time_vFB_boot_res <- get_row_median(
  data = time_vFB_t_mat,
  num_iter = 1000,
  row_num = 1)

time_vFB_violin <- create_cat_scatter(
  data = time_vFB,
  file_name = "vFB_scatter",
  plot_dir = paste0(figs_dir, "/hosts")
  )
time_vFB_violin
```
```{r vFB versus time with lifestyle}
templyt_ratio <- read.csv(
  file = paste0(data_dir, "/templyt_ratio_wide_csv.csv"),
  header = TRUE,
  row.names = 1)
templyt_ratio_t_mat <- t(templyt_ratio)

firmicute_templyt <- get_row_median(
  data = templyt_ratio_t_mat,
  num_iter = 1000,
  row_num = 2)

bacteroidota_templyt <- get_row_median(
  data = templyt_ratio_t_mat,
  num_iter = 1000,
  row_num = 1)

firmicute_templyt_violin <- create_cat_voilins(
  data = firmicute_templyt,
  ci_offset = c(0.2, 0.55, 0.55),
  file_name = "firmicute_templyt",
  plot_dir = paste0(figs_dir, "/hosts")
  )
firmicute_templyt_violin

bacteroidota_templyt_violin <- create_cat_voilins(
  data = bacteroidota_templyt,
  ci_offset = c(0.2, 0.55, 0.55),
  file_name = "bacteroidota_templyt",
  plot_dir = paste0(figs_dir, "/hosts")
  )
bacteroidota_templyt_violin
```
presence-absence heatmaps of clusters
```{r presence absense of hosts}
phage_cluster_heart <- read.csv(
  file = paste0(data_dir, "/heatmap_dataframe_heart.csv"),
  header = TRUE,
  row.names = 1
  )
create_pres_abs_heatmap(data = phage_cluster_heart,
                        lifestyle = FALSE,
                        file_name = "phage_cluster_heart",
                        plot_dir = paste0(figs_dir, "/hosts")
                        )
create_pres_abs_heatmap(data = phage_cluster_heart,
                        lifestyle = TRUE,
                        file_name = "phage_cluster_heart",
                        plot_dir = paste0(figs_dir, "/hosts")
                        )

phage_cluster_john <- read.csv(
  file = paste0(data_dir, "/heatmap_dataframe_john.csv"),
  header = TRUE,
  row.names = 1
  )
create_pres_abs_heatmap(data = phage_cluster_john,
                        lifestyle = FALSE,
                        file_name = "phage_cluster_john",
                        plot_dir = paste0(figs_dir, "/hosts")
                        )
create_pres_abs_heatmap(data = phage_cluster_john,
                        lifestyle = TRUE,
                        file_name = "phage_cluster_john",
                        plot_dir = paste0(figs_dir, "/hosts")
                        )


phage_cluster_mike <- read.csv(
  file = paste0(data_dir, "/heatmap_dataframe_mike.csv"),
  header = TRUE,
  row.names = 1
  )
create_pres_abs_heatmap(data = phage_cluster_mike,
                        lifestyle = FALSE,
                        file_name = "phage_cluster_mike",
                        plot_dir = paste0(figs_dir, "/hosts")
                        )
create_pres_abs_heatmap(data = phage_cluster_mike,
                        lifestyle = TRUE,
                        file_name = "phage_cluster_mike",
                        plot_dir = paste0(figs_dir, "/hosts")
                        )

phage_cluster_milo <- read.csv(
  file = paste0(data_dir, "/heatmap_dataframe_milo.csv"),
  header = TRUE,
  row.names = 1
  )
create_pres_abs_heatmap(data = phage_cluster_milo,
                        lifestyle = FALSE,
                        file_name = "phage_cluster_milo",
                        plot_dir = paste0(figs_dir, "/hosts")
                        )
create_pres_abs_heatmap(data = phage_cluster_milo,
                        lifestyle = TRUE,
                        file_name = "phage_cluster_milo",
                        plot_dir = paste0(figs_dir, "/hosts")
                        )

phage_cluster_nub <- read.csv(
  file = paste0(data_dir, "/heatmap_dataframe_nub.csv"),
  header = TRUE,
  row.names = 1
  )
create_pres_abs_heatmap(data = phage_cluster_nub,
                        lifestyle = FALSE,
                        file_name = "phage_cluster_nub",
                        plot_dir = paste0(figs_dir, "/hosts")
                        )
create_pres_abs_heatmap(data = phage_cluster_nub,
                        lifestyle = TRUE,
                        file_name = "phage_cluster_nub",
                        plot_dir = paste0(figs_dir, "/hosts")
                        )

phage_cluster_rob <- read.csv(
  file = paste0(data_dir, "/heatmap_dataframe_rob.csv"),
  header = TRUE,
  row.names = 1
  )
create_pres_abs_heatmap(data = phage_cluster_rob,
                        lifestyle = FALSE,
                        file_name = "phage_cluster_rob",
                        plot_dir = paste0(figs_dir, "/hosts")
                        )
create_pres_abs_heatmap(data = phage_cluster_rob,
                        lifestyle = TRUE,
                        file_name = "phage_cluster_rob",
                        plot_dir = paste0(figs_dir, "/hosts")
                        )

phage_cluster_roger <- read.csv(
  file = paste0(data_dir, "/heatmap_dataframe_roger.csv"),
  header = TRUE,
  row.names = 1
  )
create_pres_abs_heatmap(data = phage_cluster_roger,
                        lifestyle = FALSE,
                        file_name = "phage_cluster_roger",
                        plot_dir = paste0(figs_dir, "/hosts")
                        )
create_pres_abs_heatmap(data = phage_cluster_roger,
                        lifestyle = TRUE,
                        file_name = "phage_cluster_roger",
                        plot_dir = paste0(figs_dir, "/hosts")
                        )

phage_cluster_smoothie <- read.csv(
  file = paste0(data_dir, "/heatmap_dataframe_smoothie.csv"),
  header = TRUE,
  row.names = 1
  )
create_pres_abs_heatmap(data = phage_cluster_smoothie,
                        lifestyle = FALSE,
                        file_name = "phage_cluster_smoothie",
                        plot_dir = paste0(figs_dir, "/hosts")
                        )
create_pres_abs_heatmap(data = phage_cluster_smoothie,
                        lifestyle = TRUE,
                        file_name = "phage_cluster_smoothie",
                        plot_dir = paste0(figs_dir, "/hosts")
                        )

phage_cluster_tether <- read.csv(
  file = paste0(data_dir, "/heatmap_dataframe_tether.csv"),
  header = TRUE,
  row.names = 1
  )
create_pres_abs_heatmap(data = phage_cluster_tether,
                        lifestyle = FALSE,
                        file_name = "phage_cluster_tether",
                        plot_dir = paste0(figs_dir, "/hosts")
                        )
create_pres_abs_heatmap(data = phage_cluster_tether,
                        lifestyle = TRUE,
                        file_name = "phage_cluster_tether",
                        plot_dir = paste0(figs_dir, "/hosts")
                        )
```
## metabolic profile analysis
First, using the raw counts data, perform a functional diversity assessment of
the 3 sample categories.
```{r metabolic gene analysis: diversity}
# read in raw counts data
vir_ec_counts <- read.csv(
  file = paste0(data_dir, "/vir_phage_ec_counts.csv"),
  header = TRUE,
  row.names = 1
  )

# perform diversity assessment of counts
diversity_data <- get_diversity_indeces(
  data = t(vir_ec_counts),
  div_index = "shannon",
  num_iter = max_iter
)

# general plot of raw diversity numbers
diversity_plot <- create_diversity_plots(
  data = diversity_data[[1]],
  div_index = "Shannon", 
  file_name = "shannon_raw",
  plot_dir = paste0(figs_dir, "/diversity")
  )
diversity_plot

# plot of boostrapped medians
diversity_violin <- create_cat_voilins(
  data = diversity_data[[2]], 
  ci_offset = c(0.2, 0.55, 0.55),
  file_name = "shannon_medians",
  plot_dir = paste0(figs_dir, "/diversity"))
diversity_violin
```

perform an initial kmeans clustering of ec counts onto an mds
```{r metabolic gene analysis: clustering}
# normalize raw observation
vir_ec_norm <- normalize_data(df = vir_ec_counts)
vir_ec_t_norm <- t(vir_ec_norm)

# find optimal number of clusters
vir_ec_optimal <- get_optimal_clusters(
  data = vir_ec_t_norm,
  method = "silhouette"
  )
vir_ec_optimal

# generate representative mds of clutering
library(ade4)
k <- 3

kmeans_result <- kmeans(vir_ec_t_norm, centers = k)
cluster_assignments <- kmeans_result$cluster
mds <- cmdscale(dist(vir_ec_t_norm))
mds_df <- as.data.frame(mds)
mds_df$clusters <- as.factor(cluster_assignments)
mds_df$groups <- as.factor(substr(rownames(mds_df), 1, 3))
mds_df$ori <- as.factor(substr(rownames(mds_df), 5, 7))
mds_order <- data.frame(
  cat = c("pal", "ind", "pre"),
  order = c(3, 2, 1)
  )
mds_df <- merge(mds_df,  mds_order, by.x = "groups", by.y = "cat")
mds_df <- mds_df[order(mds_df$order), ]

{svglite(filename = paste0(figs_dir, "/metabolism/mds.svg"),
          width = 7,
          height = 7)
plot(mds_df$V1, mds_df$V2, pch = 19, cex = 1, col = mds_df$groups)
s.class(mds_df[, c("V1", "V2")], fac = mds_df$clusters, cell = 2, axesell = FALSE,
        csta = 0,
        col = c("red", "green", "blue"), clabel = FALSE, pch = NA,
        add.plot = TRUE)}

# plot(mds_df$V1, mds_df$V2, pch = 19, cex = 1, col = mds_df$groups,
#      xlim = c(-30, 30), ylim = c(-65, 15))
```

use bootstrap and permutation testing to determine clustering behaviour
```{r metabolic gene analysis: bootstrap and permutation}
# perform kmeans with bootstrap resampling and permutation testing
vir_ec_boot_res <- kmeans_bootstrap(
  data = vir_ec_t_norm,
  k = k,
  num_iter = max_iter
  )
vir_ec_perm_res <- kmeans_permute(
  data = vir_ec_t_norm,
  k = k,
  num_iter = max_iter
  )

# combining bootstrap and permutation results
vir_ec_clus_res <- get_clustering_res(
  boot_data = vir_ec_boot_res,
  perm_data = vir_ec_perm_res,
  star_bounds = star_bounds,
  num_iter = max_iter
  )

# plot clustering results
vir_ec_clus_violin <- create_clus_violin(
  boot_data = vir_ec_boot_res,
  clus_res = vir_ec_clus_res,
  file_name = paste0("clustering_vir_ec_", k),
  plot_dir = paste0(figs_dir, "/metabolism")
  )
vir_ec_clus_violin
```

determine which genes are most differential represented and visualize with
heatmaps and jitter plots
```{r metabolic gene analysis: differential representation}
# get differential represented genes
vir_ec_diff_repres <- get_diff_repres_cols(data = vir_ec_t_norm)
vir_ec_diff_repres <- vir_ec_diff_repres[
  order(vir_ec_diff_repres[, "kruskal_p"]), 
  ]

# generate heatmap of most differentially represented genes
vir_ec_t_scale <- scale(vir_ec_t_norm)
create_metabolic_heatmap(
  data = vir_ec_t_norm,
  signif_res = vir_ec_diff_repres,
  subset = subset_top,
  file_name = paste0("diff_repres_vir_ec_", subset_top),
  plot_dir = paste0(figs_dir, "/metabolism")
  )
create_metabolic_heatmap(
  data = vir_ec_t_scale,
  signif_res = vir_ec_diff_repres,
  subset = subset_top,
  file_name = paste0("diff_repres_vir_ec_scaled", subset_top),
  plot_dir = paste0(figs_dir, "/metabolism")
  )
create_legend(
   color_set =  brewer.pal(7, "PuRd"), 
   filename = paste0(figs_dir, "/metabolism/legend_metabolic.svg")
 )

# show ec proportions
vir_ec_median <- get_cat_medians(data = vir_ec_t_norm)

# print stats testing of ec classes
get_diff_repres_ec(data = vir_ec_median)
```

## virulence factors analysis
adjust meta data and reference data to vfdb for the rest of analysis
```{r vfdb alignment: format reference}
# read vf reference
vf_meta_data <- read.csv(
  file = paste0(data_dir, "/VFs.csv"),
  header = TRUE)
vf_reference <- vf_meta_data
colnames(vf_reference) <- vf_reference[1, ]
vf_reference <- vf_reference[-1, c("VFID", "VFCID", "VF_Name", "VFcategory")]

# read in vf counts data
vf_counts <- read.csv(
  file = paste0(data_dir, "/vf_alignment_counts.csv"),
  header = TRUE,
  row.names = 1
  )
```

determine how dense the presence of vfs are on phage contigs
```{r vfdb alignment: prevalence}
# read in vf counts data
vf_contig_hits <- read.csv(
  file = paste0(data_dir, "/vf_alignment_contig_hits.csv"),
  header = TRUE,
  row.names = 1)
vf_contig_hits["perc", ] <-
  vf_contig_hits["unique_cover", ] / vf_contig_hits["total_length", ]

# bootstrap median ratio of vf coverage to total contig lengths
vf_aln_boot_res <- get_row_median(
  data = vf_contig_hits,
  num_iter = max_iter,
  row_num = 3
  )

# plot violin plot of coverage by category
vf_aln_voilin <- create_cat_voilins(
  data = vf_aln_boot_res,
  ci_offset = c(0.15, 0.55, 0.55),
  file_name = "vf_aln",
  plot_dir = paste0(figs_dir, "/vf"))
vf_aln_voilin

# print stats testing of coverage between categories
vf_aln <- melt(vf_aln_boot_res)
kruskal.test(vf_aln$value, vf_aln$Var2)
```

determine which genes are most differential represented and determine how many
to show on the hitmap
```{r vfdb alignment: differential representation}
# normalize
vf_norm <- normalize_data(vf_counts)
vf_t_norm <- t(vf_norm)
vf_t_scale <- scale(vf_t_norm)

# get differential represented genes and combine with vf categories
vf_diff_repres <- get_diff_repres_cols(data = vf_t_norm)
vf_diff_repres <- cbind(rownames(vf_diff_repres), vf_diff_repres)
colnames(vf_diff_repres) <- c("VFID", colnames(vf_diff_repres)[-1])
vf_diff_repres <- merge(x = vf_diff_repres, y = vf_reference, by = "VFID")
vf_diff_repres$kruskal_p <- as.numeric(vf_diff_repres$kruskal_p)
vf_diff_repres$`ind vs. pre` <- as.numeric(vf_diff_repres$`ind vs. pre`)
vf_diff_repres$`ind vs. pal` <- as.numeric(vf_diff_repres$`ind vs. pal`)
vf_diff_repres$`pre vs. pal` <- as.numeric(vf_diff_repres$`pre vs. pal`)
vf_diff_repres <- vf_diff_repres[order(vf_diff_repres[, "kruskal_p"]), ]

# find most represented vfc
vfc_freq <- table(vf_diff_repres[, "VFCID"])
vfc_freq <- as.data.frame(vfc_freq)
vfc_freq <- merge(vfc_freq, unique(vf_meta_data[-1, c(5,6)]),
                  by.x = "Var1", by.y = "X.3")
vfc_freq <- vfc_freq[rev(order(vfc_freq$Freq)), ]
vfc_freq

vfc_sig_freq <- table(vf_diff_repres[vf_diff_repres$kruskal_p <= 0.05, "VFCID"])
vfc_sig_freq <- as.data.frame(vfc_sig_freq)
vfc_sig_freq <- merge(vfc_sig_freq, unique(vf_meta_data[-1, c(5,6)]),
                      by.x = "Var1", by.y = "X.3")
vfc_sig_freq <- vfc_sig_freq[rev(order(vfc_sig_freq$Freq)), ]
vfc_sig_freq

vfc_subset_freq <- table(vf_diff_repres[1:subset_top, "VFCID"])
vfc_subset_freq <- as.data.frame(vfc_subset_freq)
vfc_subset_freq <- merge(vfc_subset_freq, unique(vf_meta_data[-1, c(5,6)]), 
                         by.x = "Var1", by.y = "X.3")
vfc_subset_freq <- vfc_subset_freq[rev(order(vfc_subset_freq$Freq)), ]
vfc_subset_freq


# determine the number of vfc to use in heatmap with jenk's breaks
library(BAMMtools)
vfc_break <- getJenksBreaks(var = vfc_subset_freq$Freq, k = 3)[[2]]
vfc_colors <- which(vfc_subset_freq$Freq == vfc_break)
```

generate heatmap consistent with jenks breaks for k = 3
```{r vfdb alignment: heatmap of differential represenentation}
# generate heatmap of most differential represented genes
create_vf_heatmap(
  data = vf_t_scale,
  signif_res = vf_diff_repres, 
  subset = subset_top,
  vfc_subset = vfc_subset_freq,
  vfc_max_colors = vfc_colors,
  file_name = paste0("diff_repress_vf_profiles_", vfc_colors),
  plot_dir = heat_dir <- paste0(figs_dir, "/vf")
  )

create_legend(
   color_set = brewer.pal(vfc_colors, "PuRd"), 
   filename = paste0(figs_dir, "/vf/legend_vf_", vfc_colors, ".svg")
 )

# show ec proportions
vf_median <- get_cat_medians(data = vf_t_norm)

# print stats testing of ec classes
get_diff_repres_vf(
  data = vf_median,
  vf_reference = vf_reference
  )
```
