---
title: "Analysis of Viromes"
author: "Aydin Karatas"
output: html_document
---

## Set directory of functions
```{r setup, include=FALSE}
dir = "~/Documents/Research/project_coprolite_viromes/analysis/"
knitr::opts_chunk$set(root.dir = dir)
```

## Source Functions
```{r source functions, echo=FALSE}
base_dir <- "~/Documents/Research/project_coprolite_viromes/analysis/"

# function to source R files from a specified directory
source_files_from_dir <- function(directory) {
  full_path <- file.path(base_dir, directory)
  functions_list <- list.files(
    path = full_path,
    pattern = "*.R",
    full.names = TRUE
    )
  sapply(functions_list, source)
}

# source functions from different directories
source_files_from_dir("stats")
source_files_from_dir("graphing")
source_files_from_dir("misc")
```

## Define hyperparameters and global variables
```{r hyperpars and vars}
set.seed(seed = 123)

k <- 4
max_iter <- 1000
subset_top <- 100
color_range <- 10
star_bounds <- list(list(0.1, 0.9), list(0.025, 0.975), list(0.01, 0.99),
                    list(0.001, 0.999))

data <- "~/Documents/Research/project_coprolite_viromes_NOT_CODE/data"
vir_phage_ec_prop_data <- paste0(data, "/vir_phage_ec_counts.csv")
lifestyle_counts_data <- paste0(data, "/lifestyle_counts.csv")
cazy_prop_data <- paste0(data, "/cazy_alignment_counts.csv")
cazy_contig_hits_data <- paste0(data, "/cazy_alignment_contig_hits.csv")
vf_prop_data <- paste0(data, "/vf_alignment_counts.csv")
vf_contig_hits_data <- paste0(data, "/vf_alignment_contig_hits.csv")
vf_reference <- paste0(data, "/VFs.csv")

figures <- "~/Documents/Research/project_coprolite_viromes_NOT_CODE/figures"
violin_dir <- paste0(figures, "/violin")
heat_dir <- paste0(figures, "/heat")
jitter_dir <- paste0(figures, "/jitter")

voilin_vir_ec <- paste0("clustering_vir_ec_", k)
voilin_lifestyle <- "lifestyle"
voilin_vf_aln <- "vf_aln"
voilin_cazy_aln <- "cazy_aln"
voilin_phage_ec <- paste0("clustering_phage_ec_", k)
voilin_phage_fam <- paste0("clustering_phage_fam_", k)
voilin_phage_host <- paste0("clustering_phage_host_", k)

heat_vir_ec <- paste0("diff_repres_vir_ec_", subset_top)
heat_vir_ec_no_key <- paste0("diff_repres_vir_ec_no_key", subset_top)
heat_phage_ec <- paste0("diff_repres_phage_ec_", subset_top)
heat_phage_fam <- paste0("diff_repres_phage_fam_", subset_top)
heat_phage_host <- paste0("diff_repres_phage_host_", subset_top)

jitter_vir_classes <- paste0("vir_classes_", subset_top)
jitter_phage_classes <- paste0("phage_classes_", subset_top)
```

## Analysis of viral metabolic ec counts
First examine what the optimal k for k-means should be 
```{r vir ec prep}
# read in raw counts data
vir_ec_counts <- read.csv(
  file = vir_phage_ec_prop_data,
  header = TRUE,
  row.names = 1
  )

# normalize observations within a sample and transform to log( CPM + 1 )
vir_ec_norm <- normalize_data(df = vir_ec_counts)
vir_ec_t_mat <- scale(t(vir_ec_norm))

# find optimal number of clusters with a wss plot
vir_ec_optimal_wss <- get_optimal_clusters(
  data = vir_ec_t_mat,
  method = "wss"
  )
vir_ec_optimal_wss
```

Next, perform the bootstrap and permutation k-means
```{r vir ec boot/perm}
# perform kmeans with bootstrap resampling
vir_ec_boot_res <- kmeans_bootstrap(
  data = vir_ec_t_mat,
  k = k,
  num_iter = max_iter
  )

# perform kmeans with permutations
vir_ec_perm_res <- kmeans_permute(
  data = vir_ec_t_mat,
  k = k,
  num_iter = max_iter
  )
```

Produce a violin plot
```{r vir ec violin}
# combining bootstrap and permutation results
vir_ec_clus_res <- get_clustering_res(
  boot_data = vir_ec_boot_res,
  perm_data = vir_ec_perm_res,
  star_bounds = star_bounds,
  num_iter = max_iter
  )

# plot clustering results
vir_ec_clus_violin <- create_clus_violin(
  boot_data = vir_ec_boot_res,
  clus_res = vir_ec_clus_res,
  file_name = voilin_vir_ec,
  plot_dir = violin_dir
  )
vir_ec_clus_violin
```

Produce heatmap of most differential represented metabolic genes
```{r vir ec heat}
# get differential represented genes
vir_ec_diff_repres <- get_diff_repres_ec(data = vir_ec_t_mat)

# generate heatmap of most differential represented genes
create_heatmap(
  data = vir_ec_t_mat,
  signif_res = vir_ec_diff_repres,
  color_range = color_range,
  subset = subset_top,
  file_name = heat_vir_ec,
  plot_dir = heat_dir
  )
```

Produce jitter plot of enzyme categories
```{r vir ec ezy}
# show ec proportions
vir_ec_median <- get_cat_medians(
  data = vir_ec_t_mat[, rownames(vir_ec_diff_repres)[1:subset_top]]
  )

# print stats testing of ec classes
get_diff_repres_class(data = vir_ec_median)

# plot jitter plot
vir_ec_jitter <- create_jitter(
  data = vir_ec_median,
  classes_to_use = c(1:7),
  file_name = jitter_vir_classes,
  plot_dir = jitter_dir
  )
```

## Lifestyle prediction 
First, bootstrap lifestyle stats
```{r lifestyle boot}
# read in lifestyle counts
lifestyle_counts <- read.csv(
  file = lifestyle_counts_data,
  header = TRUE,
  row.names = 1
  )

# bootstrap median ratio of temperate-to-lytic viruses
lifestyle_boot_res <- get_median_ratios(
  data = lifestyle_counts,
  num_iter = max_iter,
  numer = 1,
  denom = 2
  )
```

Produce a violin plot
```{r lifestyle violin}
# plot violin plot of lifestyles by category
lifestyle_voilin <- create_cat_voilins(
  data = lifestyle_boot_res,
  file_name = voilin_lifestyle,
  plot_dir = violin_dir)
lifestyle_voilin

# print stats testing of lifestyles preferences between categories
lifestyle <- melt(lifestyle_boot_res)
kruskal.test(lifestyle$value, lifestyle$Var2)
```

## Alignment Results
### CAZY Alignment Counts
```{r cazy alignment}
# read in cazy counts data
cazy_counts <- read.csv(
  file = cazy_prop_data,
  header = TRUE,
  row.names = 1
  )

#TODO: format cazy counts

```

### CAZY Contig Hits
First, bootstrap median coverage of db proteins
```{r cazy contig hits boot}
# read in cazy counts data
cazy_contig_hits <- read.csv(
  file = cazy_contig_hits_data, 
  header = TRUE,
  row.names = 1)

# bootstrap median ratio of cazy coverage-to-total contig lengths
cazy_aln_boot_res <- get_median_ratios(
  data = cazy_contig_hits,
  num_iter = max_iter,
  numer = 2,
  denom = 1)
```

Produce violin plot.
```{r cazy contig hits violin}
# plot violin plot of coverage by category
cazy_aln_voilin <- create_cat_voilins(
  data = cazy_aln_boot_res,
  file_name = voilin_cazy_aln,
  plot_dir = violin_dir)
cazy_aln_voilin

# print stats testing of coverage between categories
cazy_aln <- melt(cazy_aln_boot_res)
kruskal.test(cazy_aln$value, cazy_aln$Var2)
```

### VFDB Alignment
```{r vfdb alignment}
# read vf reference
vf_to_vfc <- read.csv(
  file = vf_reference,
  header = TRUE)

# format vf reference
colnames(vf_to_vfc) <- vf_to_vfc[1, ]
vf_to_vfc <- vf_to_vfc[-1, c("VFID", "VFCID")]

# read in vf counts data
vf_counts <- read.csv(
  file = vf_prop_data,
  header = TRUE,
  row.names = 1
  )

#TODO: format vf counts
vfc_counts <- format_vf_counts(
  data = vf_counts,
  vf_ref = vf_to_vfc
)
```

### VFDB Contig Hits
First, bootstrap median coverage of db proteins
```{r vf contig htis}
# read in vf counts data
vf_contig_hits <- read.csv(
  file = vf_contig_hits_data,
  header = TRUE,
  row.names = 1)

# bootstrap median ratio of vf coverage to total contig lengths
vf_aln_boot_res <- get_median_ratios(
  data = vf_contig_hits,
  num_iter = max_iter,
  numer = 2,
  denom = 1
  )
```

Produce violin plots
```{r vf contig hits violin}
# plot violin plot of coverage by category
vf_aln_voilin <- create_cat_voilins(
  data = vf_aln_boot_res,
  file_name = voilin_vf_aln,
  plot_dir = violin_dir)
vf_aln_voilin

# print stats testing of coverage between categories
vf_aln <- melt(vf_aln_boot_res)
kruskal.test(vf_aln$value, vf_aln$Var2)
```



